---
title: 'Red "Vinho Verde" Wine Quality Exploration'
author: 'Leonardo Rota-Rossi'
output: 
  html_document:
    toc: TRUE
---


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
winedata = read.csv('./wineQualityReds.csv')
```

# Introduction

### Data Source[^1]

The dataset is related to red variant of the Portuguese "Vinho Verde" wine.
For more details, consult: http://www.vinhoverde.pt/en/.
Due to privacy and logistic issues, only physicochemical (inputs) and sensory 
(the output) variables are available (e.g. there is no data about grape types, 
wine brand, wine selling price, etc.).

The inputs include objective tests (e.g. pH values) and the output is based on 
  sensory data (median of at least 3 evaluations made by wine experts). Each 
  expert graded the wine quality between 0 (very bad) and 10 (very excellent).

#### Attribute information

- Number of items: `r nrow(winedata)`.
- Number of attributes: `r ncol(winedata)`.

##### Input variables (based on physicochemical tests)

1. fixed acidity (tartaric acid - g/dm^3^)
2. volatile acidity (acetic acid - g/dm^3^)
3. citric acid (g/dm^3^)
4. residual sugar (g/dm^3^)
5. chlorides (sodium chloride - g/dm^3^)
6. free sulfur dioxide (mg/dm^3^)
7. total sulfur dioxide (mg/dm^3^)
8. density (g/cm^3^)
9. pH
10. sulphates (potassium sulphate - g/dm^3^)
11. alcohol (% by volume)

#### Description of attributes

1. fixed.acidity: most acids involved with wine or fixed or nonvolatile 
  (do not evaporate readily);
2. volatile.acidity: the amount of acetic acid in wine, which at too high of 
  levels can lead to an unpleasant, vinegar taste;
3. citric.acid: found in small quantities, citric acid can add 'freshness' and 
  flavor to wines;
4. residual.sugar: the amount of sugar remaining after fermentation stops, it's
  rare to find wines with less than 1 gram/liter and wines with greater than 45 
  grams/liter are considered sweet;
5. chlorides: the amount of salt in the wine;
6. free.sulfur.dioxide: the free form of SO~2~ exists in equilibrium between 
  molecular SO~2~ (as a dissolved gas) and bisulfite ion; it prevents microbial
  growth and the oxidation of wine;
7. total.sulfur.dioxide: amount of free and bound forms of SO~2~; in low 
  concentrations, SO~2~ is mostly undetectable in wine, but at free SO~2~
  concentrations over 50 ppm, SO~2~ becomes evident in the nose and taste of 
  wine;
8. density: the density of water is close to that of water depending on the
  percent alcohol and sugar content;
9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic)
  to 14 (very basic); most wines are between 3-4 on the pH scale;
10. sulphates: a wine additive which can contribute to sulfur dioxide gas 
  (SO~2~) levels, wich acts as an antimicrobial and antioxidant;
11. alcohol: the percent alcohol content of the wine;

##### Output variable (based on sensory data)

12. quality (score between 0 and 10).

------

# Descriptive Statistics

```{r echo=FALSE, Descriptive_Statistics1}
  summary(winedata)
```

Quality grades distribution: 

```{r echo=FALSE, Descriptive_Statistics2}
  table(winedata$quality)
```

As we can see on the statistics summary above, no wines have quality value 
smaller than 3 or bigger than 8. Also, we can see that the quality value 
is discrete, and should be treated as an ordinal variable.

```{r echo=FALSE, Descriptive_Statistics3}
winedata$quality <- factor(winedata$quality)
```

------

# Univariate Plots Section

Let's start by looking on what are the distributions by some of the variables:

```{r echo=FALSE, Univariate_Plots1}

theme_set(theme_minimal(10))

plot1 <- ggplot(winedata, aes(fixed.acidity)) +
    geom_histogram(binwidth = 1) +
    xlab('g/l') +
    ggtitle('Wines by Fixed Acidity')

plot2 <- ggplot(winedata, aes(volatile.acidity)) +
    geom_histogram(binwidth=0.05) +
    xlab('g/l') +
    ggtitle('Wines by Volatile Acidity')

plot3 <- ggplot(winedata, aes(citric.acid)) +
    geom_histogram(binwidth=0.05) +
    xlab('g/l') +
    ggtitle('Wines by Citric Acid')

plot4 <- ggplot(winedata, aes(residual.sugar)) +
    geom_histogram(binwidth=1) +
    xlab('Residual sugar(g/l)') +
    ggtitle('Wines by Residual Sugar')

plot5 <- ggplot(winedata, aes(chlorides)) +
    geom_histogram(binwidth=0.01) +
    xlab('Chlorides (g/l)') +
    ggtitle('Wines by Chlorides')

plot6 <- ggplot(winedata, aes(density)) +
    geom_histogram(binwidth = 0.001) +
    xlab('g/ml') +
    ggtitle('Wines by Density')

plot7 <- ggplot(winedata, aes(pH)) + 
  geom_histogram(binwidth = 0.05) +
  ggtitle('Wines by pH Level')

plot8 <- ggplot(winedata, aes(alcohol)) + 
  geom_histogram(binwidth = 0.1) +
  xlab('Percentage') +
  ggtitle('Wines by Alcohol %')

plot9 <- ggplot(data=winedata, aes(x=quality)) + 
  geom_bar() + 
  xlab('Quality') +
  ggtitle('Wines by Quality')


grid.arrange(plot1, plot2, plot3, plot4, plot5, plot6, plot7, plot8, plot9, ncol=3)
```

Taking a first look on the graphs, we can see that the quality has a somewhat
normal distribution. The same happens with pH and Density. the distribution is 
mostly right skewed for all other attributes, which seems to point to consistent 
low concentrations of those attributes.

Specifically, let's take a look on the Sulfites and Sulphates distributions:

```{r echo=FALSE, Univariate_Plots2}
plot10 <- ggplot(winedata, aes(sulphates)) +
    geom_histogram(binwidth=0.05) +
    xlab('g/l') +
    ggtitle('Wines by Sulphates')

plot11 <- ggplot(winedata, aes(free.sulfur.dioxide)) +
    geom_histogram(binwidth = 2) +
    xlab('mg/l') +
    ggtitle('Wines by Sulfur Dioxide Free')

plot12 <- ggplot(winedata, aes(total.sulfur.dioxide)) +
    geom_histogram(binwidth = 10) +
    xlab('mg/l') +
    ggtitle('Wines by Sulfur Dioxide Total')

grid.arrange(plot10, plot11, plot12, ncol=3)

```

As we can see, most wines have a low concentration of those compounds, with just
few of them having a higher amount of sulphates and sulfites.

------

# Univariate Analysis

### What is the structure of your dataset?

There are 1599 wines in the dataset with 12 features (as seen above). 
The variable quality is discrete and varies from 0 to 10, but in this dataset, 
the minimum is 3 and the maximum 8.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest in this dataset is the quality. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The other features will be used to investigate their influence in the perceived
quality of the wine, in special the ones that relate to the perceived flavor 
(like `volatile acidity`, `residual sugar` and `chloride`).

The presence of sulphates and SO~2~ (sulfites) is also evaluated. Sulfites are 
generated by the fermentation and aging processes, and may taint the wine 
flavor. One common way to balance this effect is to add sulphate - usually 
Copper Sulfate (CuSO~4~) to reduce the formation of sulfites. I will investigate
how the levels of sulfites and sulphates affect the perceived quality of the
wine.

------

# Bivariate Plots Section

```{r echo=FALSE, Bivariable_Plots1}
ggplot(winedata, aes(alcohol, density)) +
  geom_point(alpha=0.2) +
  ggtitle('Density by Alcohol Percentage') +
  labs(x='Alcohol (% by volume)', y=expression('Density (g/cm'^3*')')) +
  geom_smooth()
```

Correlation between Density and Alcohol level:

```{r echo=FALSE}
with(winedata, cor.test(density, alcohol))
```

Here, we start by evaluating how the alcohol percentage affects the density. 
The graph and the correlation index are consistent with: 
- Beverages are mostly water;
- Water is more dense than alcohol;
- Higher percentages of alcohol make a wine less dense.

```{r echo=FALSE, Bivariate_Plots2}
ggplot(data=winedata, aes(x=quality, y=alcohol)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=0.2, color='orange') +
  geom_point(stat = 'summary', fun.y = mean, shape=4 ) +
  labs(x='Quality', y='Alcohol (% by volume)') +
  ggtitle('Quality by Alcohol level')
```

Correlation between Quality and Alcohol level:

```{r echo=FALSE}
with(winedata, cor.test(as.numeric(quality), alcohol))
```

We can see that the best evaluated wines have consistent higher levels of 
alcohol, and also the presence of a high number of outliers in the quality 
grade 5. The correlation test seems to point in this direction too.

```{r echo=FALSE, Bivariate_Plots3}
ggplot(data=winedata, aes(x=quality, y=pH)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=0.2, color='orange') +
  geom_point(stat = 'summary', fun.y = mean, shape=4 ) +
  labs(x='Quality', y='pH Level') +
  ggtitle('Quality by pH level')
```

Correlation between Quality and pH level:

```{r echo=FALSE}
with(winedata, cor.test(as.numeric(quality), pH))
```

We can see that the pH level has null to very little effect on the perceived
quality of the wine.


```{r echo=FALSE, Bivariable_Plots4}
ggplot(winedata, aes(x=quality, y=citric.acid)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=0.2, color='orange') +
  geom_point(stat = 'summary', fun.y = mean, shape=4 ) +
  labs(x='Quality', y='Citric Acid (g/l)') +
  ggtitle('Quality by Acid Citric Level')
```

Correlation between Quality and Citric Acid level:

```{r echo=FALSE}
with(winedata, cor.test(as.numeric(quality), citric.acid))
```

The graph and the calculations seems to indicate a small correlation between 
the Citric Acid amount and the perceived quality of the wine.

```{r echo=FALSE, Bivariate_Plots5}
ggplot(data=winedata, aes(x=quality, y=residual.sugar)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=1/10) +
  labs(x='Quality', y='Residual Sugar (g/l)') +
  ggtitle('Quality by Residual Sugar level')
```

Correlation between Quality and Residual Sugar level:

```{r echo=FALSE}
with(winedata, cor.test(as.numeric(quality), residual.sugar))
```

The quality also seems to not be affected by the residual sugar, but there are
several outliers in this case.

```{r echo=FALSE, Bivariate_Plots6}
ggplot(data=winedata, aes(x=quality, y=sulphates)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=1/5, color='lightblue') +
  labs(x='Quality', y='Sulphates (g/l)') +
  ggtitle('Quality by Sulphates level') 
```

Correlation between Quality and Sulphates level:

```{r echo=FALSE}
with(winedata, cor.test(as.numeric(quality), sulphates))
```

The quality seems to be lightly affected by the presence of sulphates, and there
are several outliers in the 5-6 quality range.

```{r echo=FALSE, Bivariate_Plots7}
ggplot(data=winedata, aes(x=quality, y=volatile.acidity)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=1/5, color='pink') +
  geom_point(stat = 'summary', fun.y = mean, shape=4) +
  labs(x='Quality', y='Volatile Acidity (g/l)') +
  ggtitle('Quality by Volatile Acidity level')
```

Correlation between Quality and Volatile Acidity level:

```{r echo=FALSE}
with(winedata, cor.test(as.numeric(quality), volatile.acidity))
```

The volatile acidity is negatively correlated to the quality - the less, the 
better.

------

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Our main investigation was about how the quality is affected by several 
attributes in the dataset. As can be seen above, some variables affect the 
perceived quality positively (`alcohol` or `sulphates`), some negatively
(`volatile acidity`) and some seem to not affect whatsoever (`residual sugar`).

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

The relationship between the density and the alcohol level seems strong - the 
more alcohol, less dense is the wine - which makes sense: as any beverage, wines
are mostly water. As alcohol is less dense than water, the more alcohol, less
total density.

### What was the strongest relationship you found?

The relationship between density and alcohol percentage. And among the ones
studying the wine quality, the relationship between the `alcohol` percentage and 
the perceived `quality`.

------

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots1}
ggplot(aes(y = total.sulfur.dioxide, x = free.sulfur.dioxide), data = winedata) + 
  geom_point(alpha=.4, size = winedata$sulphates*3, aes(colour=winedata$sulphates))+
  scale_colour_gradient(low='orange', high='navyblue', name='Sulphates (g/l)') +
  ggtitle(expression('Total SO'[2]*' and Free SO'[2]*' Relationship')) +
  labs(x=expression('Free SO'[2]*' (mg/l)'), y=expression('Total SO'[2]*' (mg/l)')) +
  geom_smooth(method='loess')
```


The amount of Free SO~2~ is consistent with the amount of Total SO~2~ in the 
studied wines. The number of wines showing high level of sulphates is small.

```{r echo=FALSE, Multivariate_Plots2}
ggplot(aes(y = total.sulfur.dioxide, x = sulphates), data = winedata) + 
  geom_point(alpha=.2, aes(colour=winedata$quality))+
  scale_colour_brewer("Quality", palette='Reds') +
  ggtitle(expression('Total SO'[2]*' and Sulphates Relationship with Quality')) +
  labs(x='Sulphates (g/l)', y=expression('Total SO'[2]*' (mg/l)')) +
  geom_smooth(method='lm')
```

Consistent with the previous graph, we can see that most wine have small 
concentrations of Sulphates, and that does not affect considerably the total
SO~2~.

```{r echo=FALSE, Multivariate_Plots3}
ggplot(aes(y=citric.acid, x=pH), data=winedata) +
  geom_point(aes(colour=quality)) +
  scale_colour_brewer("Quality", palette="Greens") +
  ggtitle('Citric Acid and pH levels relationship with Quality') +
  labs(x='pH level', y='Citric Acid (g/l)') +
  geom_smooth(method='lm') +
  coord_cartesian(ylim=c(0,1))
```


We can see a strong correlation between the pH level and the amount of citric
acid in the wine. The pH level seems to affect less the perceived quality than
the citric acid, however.

------

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

As we can see, most wines have low level of both Free and total SO~2~, and seem 
to use small amounts of Sulphates. Few wines have high levels of sulphates, and
that may indicate a good control over aging process by the producers.

Also, the expected relation between pH and citric acid is present (lower pH =
higher acidity). And we can see that the quality of the wine is affected by the
level of Citric Acid, but not so much by the pH level.

### Were there any interesting or surprising interactions between features?
I was expecting to see smalles levels of added sulphates on higher quality 
wines, which could indicate a better or more traditional aging process. It seems
the opposite - higher quality wines have higher sulphate levels.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(y=citric.acid, x=pH), data=winedata) +
  geom_point(alpha=1/2, aes(colour=quality), size=2.5) +
  scale_color_brewer(name='Quality') +
  ggtitle('Citric Acid and pH levels relationship with Quality') +
  labs(x='pH level', y='Citric Acid (g/l)') +
  geom_smooth(method='lm') +
  coord_cartesian(ylim=c(0,1))
```

#### Description One

As explained in the dataset, we can see that there's a light correlation between
the perceived quality and the citric acid level. Interestingly, we can see that
this relationship does not extend to the pH level - not all wines with high
citric acid level have low pH.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data=winedata, aes(x=quality, y=residual.sugar)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=1/5, color='lightblue') +
  geom_point(stat = 'summary', fun.y = mean, shape=4) +
  labs(x='Quality', y='Residual Sugar (g/l)') +
  ggtitle('Quality by Residual Sugar level')
```

#### Description Two

The residual sugar seems to not affect the perceived quality. This is somewhat 
interesting, because the *vinho verde* is not a sweet wine (which could have 
made the sweeter ones to have a poor evaluation).

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data=winedata, aes(x=quality, y=alcohol)) +
  geom_boxplot(aes(group=quality)) +
  geom_jitter(alpha=1/5, color='lightblue') +
  geom_point(stat = 'summary', fun.y = mean, shape=4) +
  labs(x='Quality', y='Alcohol (% by volume)') +
  ggtitle('Quality by Alcohol level') 
```

#### Description Three

Of all the parameters evaluated, this one seems interesting to me - the 
correlation between quality and alcohol level. There's a somewhat strong 
correlation between the alcohol level and the perceived quality of the wine. 
Maybe the reviewers were more interested in the effects of wine than flavors? :)

------

# Reflection

The redwine dataset contains 1599 observations, across 12 variables[^2], from 
sometime around 2009 for the red *vinho verde* wine. My initial approach was to
look at the variables names and their summary statistics, to identify 
interesting values for study.

My main focus was the quality variable - it is defined as the median from three 
separate reviews from expert in wines. I tried to investigate what chemical 
characteristics were consistent with the grades. 

The main challenge I found was understanding the relationship about some 
chemical processes used in the wine production - the addition of Sulphates to 
improve the aging process, for instance. Also, I was under the impression that
sweeter wines would be worse evaluated than drier ones - and the data does not
support this point of view.

Several factors seem to affect the perceived quality - some of them positively,
some not. Among the positive ones we can see that citric acid and alcohol were 
the most proeminent ones, and the volatile acidity negatively affects quality.

Some limitations need to be considered: the reviews were made by a small set of
reviewers, there are no information about the methodology adopted in these
reviews, and the dataset only studies an specific kind of wine - the red wine 
variety of *vinho verde*, without considering the specificities among them 
type of grape, for instance). 


[^1]: P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.
Modeling wine preferences by data mining from physicochemical properties.  
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.  
Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016  
  [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib  
  
[^2]: the first variable in the dataset (X), is just a sequential ID for each
observation, and was ignored.